How to make an abridged life table and calculate life expectancy

### 2016-09-20 working_teaching_examples/how_to_abridged_life_table.Rmd

Now, $L_x = n_x*(l_x - d_x) + n_x\bar{a_x}d_x$
$L_x = n_xl_x - n_xd_x + n_x\bar{a_x}d_x$
$l_x = \frac{L_x + n_xd_x - n_x\bar{a_x}d_x}{n_x}$
$l_x = \frac{L_x + d_xn_x(1 - \bar{a_x}){n_x}$

@@

$q_x=\frac{d_xn_x}{L_x + d_xn_x(1 - \bar{a_x}}$, and dividing numerator and denominator by L_x, we get
$q_x=\frac{R_xn_x}{1 + R_xn_x(1 - \bar{a_x}}$.

@@

head(canada.men)

quebec.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90), 
                         "obs.deaths" = c(1057, 164, 116, 176, 659, 956, 1056, 1110, 1498, 2470, 4116, 6054, 8426,
                                          11175, 13558, 16567, 21349, 22396, 16851,10878),
                         "obs.popn" = c(215200, 789710, 1008266, 1179423, 1249360, 1261780, 1349857, 1293521,
                                        1313834, 1551684, 1622835, 1485249, 1288014, 1077253, 796557, 609341, 
                                        476211, 300506, 134353, 49678))
head(quebec.men)
```

@@

quebec.men$n_x <- 5
quebec.men$n_x[quebec.men$ages == 0] <- 1
quebec.men$n_x[quebec.men$ages == 1] <- 4
```

```{r function_lifeexpectancy}
#dataset needs to be in order of age.
life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){

  data["R_x"] <- data[death.counts]/data[population.counts] #mortality rates

  if(is.na(ave.prop.lived) == T) {
      data["a_x"] <- 0.5 #average proportion of interval lived
      data[1, "a_x"] <- 0.09
  } else{
      data["a_x"] <- data[ave.prop.lived]
  }

  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) 
  #probability of dying

  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group

  data["p_x"] <- 1 - data["q_x"] #probability of survival

  data["l_x"] <- NA #l_x: is the number of individuals alive at the start of the interval
  data[1, "l_x"] <- 100000

  for(i in 2:dim(data)[1]) {
    data[i, "l_x"] <- data[i - 1, "l_x"]*data[i-1, "p_x"]
  }

  data["d_x"] <- data["l_x"]*data["q_x"] #number of deaths
  data["L_x"] <- (data["l_x"] - data["d_x"])*data[num.ages.in.group] + data["d_x"]*data["a_x"]*data[num.ages.in.group] #person time in the age bracket
  data[dim(data)[1], "L_x"] <- data[dim(data)[1], "l_x"]/data[dim(data)[1], "R_x"] #different for last age bracket

  #now to calculate the total person time lived at or after age x, T_x:
  data["cumsum"] <- cumsum(data["L_x"])
  data["l.cumsum"] <- c(0, unlist(data[1:(dim(data)[1] - 1), "cumsum"]))
  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
  data["e_x"] <- data["T_x"]/data["l_x"]

  return(data)

} 
```

@@

@@

life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn")

life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived")
```

```{r}
load("Data/alabama_only.Rdata")

#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))

@@

system.time(first.strata.life.table <- life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))

View(first.strata.life.table)

system.time(
for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
  assign(paste0("lt.", as.character(stratum.id)), 
         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
                                  age.groups = "Age2", num.ages.in.group = "nx", 
                                  death.counts = "total_deaths5", population.counts = "Population"))
  print(stratum.id)
}
)
```

>>>>

Now, $L_x = n_x(l_x - d_x) + n_x\bar{a_x}d_x$

$L_x = n_xl_x - n_xd_x + n_x\bar{a_x}d_x$

$l_x = \frac{L_x + n_xd_x - n_x\bar{a_x}d_x}{n_x}$

$l_x = \frac{L_x + d_xn_x(1 - \bar{a_x})}{n_x}$

@@

$q_x=\frac{d_xn_x}{L_x + d_xn_x(1 - \bar{a_x})}$, and dividing numerator and denominator by $L_x$, we get

$q_x=\frac{R_xn_x}{1 + R_xn_x(1 - \bar{a_x})}$.

@@

quebec.men <- data.frame("ages" = c(0, 1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90), 
                         "obs.deaths" = c(1057, 164, 116, 176, 659, 956, 1056, 1110, 1498, 2470, 4116, 6054, 8426,
                                          11175, 13558, 16567, 21349, 22396, 16851,10878),
                         "obs.popn" = c(215200, 789710, 1008266, 1179423, 1249360, 1261780, 1349857, 1293521,
                                        1313834, 1551684, 1622835, 1485249, 1288014, 1077253, 796557, 609341, 
                                        476211, 300506, 134353, 49678))
```

@@

quebec.men$n_x <- 5
quebec.men$n_x[quebec.men$ages == 0] <- 1
quebec.men$n_x[quebec.men$ages == 1] <- 4

head(canada.men)
head(quebec.men)
```

This source file contains the life expectancy functions used in this project:
```{r load_functions}
source("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Code/life_expectancy_functions.R")
```

@@

```{r calc_age_decomp}
age_decomp <- le_age_decomp(life.table.qm, life.table.cm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")

age_decomp2 <- le_age_decomp(life.table.cm, life.table.qm, num.alive="l_x", interval.years.lived="L_x", accumulated.years.lived = "T_x")
```

@@

data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn"))

data.frame(life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived"))
```

```{r}
#load("Data/alabama_only.Rdata")
load("/Users/corinneriddell/Documents/repos/BlackWhiteGap/Data/alabama_only.Rdata")
#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))

@@

system.time(first.strata.life.table <- data.frame(life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population")))

first.strata.life.table[, c("Age2", "total_deaths5", "Population", "R_x", "q_x", "l_x", "d_x", "L_x", "T_x", "e_x")]

system.time(
for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
  assign(paste0("lt.", as.character(stratum.id)), 
         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
                                  age.groups = "Age2", num.ages.in.group = "nx", 
                                  death.counts = "total_deaths5", population.counts = "Population"))
}
)
```
